Kelp morphometric properties and the link to hydrodynamic modelling

### 2018-10-02 chapter_1.Rmd
minor edits
added section on wave theory to chapter 1. More to come.

The important environmental drivers of kelp individuals and communities include light, substrata, salinity, sedimentation, nitrients, temperature and wave exposure. Although studies have investigated the effects of important environmental drivers, the roles these factors play is often difficult to evaluate as such factors may never be fully independent of each other, i.e. environmental factors are to some extent dependent on one another. Multifactorial studies have attempted to explain combined affects, however these studies are often limited to investigating combination of two or three environmental drivers as inclusion of too many factors can lead to results that are difficult to interept. Environmetal factors are highly variable on temporal and spatial scales and their effects may also be dependent on the life-stage of the organism, adding a further layer of complexity to investigations.

@@

>>>>

The important environmental drivers of kelp individuals and communities include light, substrata, salinity, sedimentation, nitrients, temperature and wave exposure. Although studies have investigated the effects of important environmental drivers, the roles these factors play is often difficult to evaluate as such factors may never be fully independent of each other, i.e. environmental factors are to some extent are dependent on one another. Multifactorial studies have attempted to explain combined affects, however these studies are often limited to investigating combination of two or three environmental drivers as inclusion of too many factors can lead to results that are difficult to interept. Environmetal factors are highly variable on temporal and spatial scales and their effects may also be dependent on the life-stage of the organism, adding a further layer of complexity to investigations.

@@

# Ocean and coastal waves

Regardless of the location around the world, waves are a feature of any coastline and the marine environment and are important manifestations of energy in the ocean. Waves are not the movement of water particles but is rather the movement and propagation of energy through the ocean. The source of this energy can be formed locally, such as wind-driven waves, and or from distant locations in the ocean, such as storms. This energy is then transferred from deeper water into shallower water where it plays a role in driving beach morphology. Sand and rock are eroded with each passing wave, thereby playing a role in shaping coastlines. Energy that is left-over after these processes is transferred into heat energy, and heats up the sand and rocks on which each wave hits. This energy can also be harnessed in simple ways such as a surfer catching a wave, or more complex ways such as capturing energy from the ocean environment for electricity production. However, waves are stochastic in nature and therefore the energy that propagates through the ocean is not always consistent. Therefore, it is often difficult to quantify and predict wave energy in the marine environment. Despite this fact, waves are gaining more recognition for the role they play in shaping coastlines, beaches and hence the communities and organisms that depend on these systems.

Waves manifest themselves in different ways, which is also dependent on the energy creating force, and can be classified into different categories. For instance, "chop" are produced by local winds, while "tsunamis"" are rare waves that are formed during a earthquakes or landslides and can be produced from thousands of kilometers away. Waves in the ocean are known as "swell" and are produced from far distant storms, while the most consistent form of waves that interacts with the coastlines, usually twice a day, are called "tides". Tides are produced from differences in gravitational forces between the ocean, earth's crust, sun and the moon. These classifications can be broken down further and will be investigated later in this chapter. The focus of this chapter will be understanding wave theory and how waves form and propogate into shallow-coastal waters. This will be essential to later chapters that aim to model wave energy and callibrate the produced model with biotic parameters.

## Generating and restoring forces

Waves are formed due to the constant interaction bewteen two forces, which are know as the "generating source" and the "restoration force". The generating force is the force which pushes water from one layer up into the other, for example pushing water up across the boundry of air. This can also occur in the ocean layers. The ocean layers through differences in desnsity, which in turn is partly driven by abiotic parameters such as temperature and salinity, which creates a boundry for wave energy to move along. Examples of such boundries are along the ocean surface, pycnoclines, and the sea-floor. **optional addition of pycnocline details**. In other words, waves occur along the boundries formed in the ocean by various abiotic processes, and therefore density boundries are essential to the propagation of energy through the ocean.

The generating force and restoring force occur along these boundries, the generating force pushes water up across ocean boundries while the restoring force pulls it back to where the boundry was orginally, trying to restore the balance in energy. This "tug of war" between the two forces creates an oscillating motion between the boundry layers which acts as a point of disturbance, sending out energy in all directions. The disturbance energy will continue to manifest itself provided the tug and pull between the two forces is occuring. Once the generating force stops, and ultimately the energy from the point of disturbance, the waves dissipate and water restores to its orginal state. A simlple example would be blowing on water in a cup. The blowing of air onto the surface of the water in the cup creates a point of disturbance that pushes the air boundry into the water boundry. The restoring force is the surface tension of the water that is maintained through hydrogen bonds between water molecules. Once blowing on the water in the cup has stopped, the ripples or "waves" in the cup begin to diminish in size until the surface tension returns to its orginal state. The hydrogen bonds between the water molecules are stronger than the force of gravity, and therefore the force of the surface tension returns the water to its orginal state. These waves are know as "capillary waves" and is essentially residual energy after the generating source has stopped. The restoring force can also be in the form of gravity and are know as "gravity waves". Using the same example, if one blows too hard, the water boundry is pushed up into the air boundry, breaking the hydrogen bonds between water molecules which allows gravity to return the water to its orginal state. All waves in the ocean are either capillary waves or gravity waves, and their classification will be dependent on the restoring force involved.

In nature, there are three kinds of generating forces. These are wind, displacement of large volumes of water and uneven forces of gravitational attraction bewtween the Earth, Moon and Sun. Different generating forces are associated with different wave heights, periods and the type of wave produced **insert appropriate figure**. Wind-generated waves comprise of capillary waves, chop, swell and seiche. Some seiche can form from landslides and earthquakes but most of these waves are know as a tsunami. Swell has create the waves with the large heights (up to 100m), while tides can create the tallest.

# Wave physics and scales

Waves have a number of charateristics which are dipected in figure ??, and is an idealised representation of what a wave is. The *amplitude* is the vertical distance from its midline or *equilibrium surface* or *still water level* to its highest point known as the *crest*. The *trough* is the same distance of the amplitude, however the measurement is taken from the equilibrium surface to the lowest point.
